---
title: "Incorporating external data"
output:
  html_document:
    toc: true
    theme: null
vignette: >
  %\VignetteIndexEntry{External data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

First, invent some data to use later:

```{r}
logbook_fo <- expand.grid(month = 1:12, year = 2000:2006)
logbook_fo$vessel_nr <- sample(1:100, nrow(logbook_fo), replace = TRUE)
logbook_fo$mfdb_gear_code <- "LLN"
logbook_fo$lat <- round(rnorm(nrow(logbook_fo), 61.0), 3)
logbook_fo$lon <- round(rnorm(nrow(logbook_fo), -17.5), 3)
logbook_fo$catch <- runif(nrow(logbook_fo), 1e4, 1e6)
logbook_fo$tow_area <- rnorm(nrow(logbook_fo), 12, 0.6)
logbook_fo$tow_time <- rnorm(nrow(logbook_fo), 300, 10)

logbook_gl <- expand.grid(month = 1:12, year = 2000:2006)
logbook_gl$vessel_nr <- sample(1:100, nrow(logbook_gl), replace = TRUE)
logbook_gl$mfdb_gear_code <- "LLN"
logbook_gl$lat <- round(rnorm(nrow(logbook_gl), 61.0), 3)
logbook_gl$lon <- round(rnorm(nrow(logbook_gl), -30.5), 3)
logbook_gl$catch <- runif(nrow(logbook_gl), 1e4, 1e6)
logbook_gl$tow_area <- rnorm(nrow(logbook_gl), 12, 0.6)
logbook_gl$tow_time <- rnorm(nrow(logbook_gl), 300, 10)
```

Data can be imported into a Pax DB using ``pax_import()``:

```{r}
library(pax)

# NB: In real-life this will probably be generated using pax::pax_from_mar()
pcon <- pax::pax_connect()
pax_import(pcon, pax_marmap_ocean_depth())
for (s in pax_def_strata_list()) {
  pax_import(pcon, pax_def_strata(s))
}

pax_import(pcon, logbook_fo, cite = "logbook_fo.csv")
pax_import(pcon, logbook_gl, cite = "logbook_gl.csv")
```

We can list the contents:

```{r}
pax_contents(pcon)
```

By default ``pax_import()`` will use the incoming name as the table name.
This can be overridden with the ``name`` parameter.

We can use the imported tables as we would the mar "logbook" table:

```{r}
dplyr::tbl(pcon, "logbook_gl") |>
   pax_landings_fishingyear_summary()

dplyr::tbl(pcon, "logbook_fo") |>
   pax_landings_fishingyear_summary()

dplyr::tbl(pcon, "logbook_gl") |>
   dplyr::union_all(dplyr::tbl(pcon, "logbook_fo")) |>
   pax_landings_fishingyear_summary()
```

```{r}
# i.e. hr_catch_by_location()
catch_by_location <-
  dplyr::tbl(pcon, "logbook_gl") |> dplyr::union_all(dplyr::tbl(pcon, "logbook_fo")) |>
    dplyr::group_by(year, lat = round(lat, 1), lon = round(lon, 1)) |>
    dplyr::summarise(
      catch = sum(1e-3 * catch / tow_area, na.rm = TRUE),
      tow_time = sum(tow_time / tow_area, na.rm = TRUE)
    ) |>
    dplyr::ungroup()

# i.e hr_catch_dist_plot()
  pax::pax_map_base(plot_greenland = TRUE, plot_faroes = TRUE, xlim = c(-55, 0), ylim = c(60, 67.25)) |>
    pax::pax_map_layer_depth(dplyr::tbl(pcon, "ocean_depth")) |>
    pax::pax_map_layer_catch(
      data = catch_by_location,
      alpha = 1,
      na.fill = -50,
      breaks = c(0, 1, 2, seq(3, 20, by = 3), 40, 60)
    )
```

We can also add extra ``pax_import()`` calls into a targets pipeline.

Before importing, the CSV file (assuming that is your data source) should be registered with targets.
This will invalidate the pipeline steps automatically whenever the file changes.
Note that the ``logbook_fo`` target is just the path to the CSV file, ``pax_import()`` will automatically read it in.

```{r}
library(targets)
tar_dir(local({ # tar_dir() runs code from a temp dir

  # Dump to file, assume that data will be stored alongside model
  write.csv(logbook_fo, file = "logbook_fo.csv")

  # Write _targets.R
  tar_script({
    library(targets)
    library(pax)
    list(
      # Register CSV file in targets 
      tar_target(logbook_fo, "logbook_fo.csv", format = "file"),
      # Populate local database with standard data and extras
      tar_target(pax_db, {
        # NB: Ideally pcon <- pax_from_mar(species, year_start, year_end)
        pcon <- pax::pax_connect()
        pax_import(pcon, logbook_fo, cite = "logbook_fo.csv")
        pcon
      }, format = pax_tar_format_duckdb()),
      # Save contents of DB
      tar_target(db_contents, pax_contents(pax_db), format = pax::pax_tar_format_parquet())
    )
  })

  tar_make()
  print(tar_read(db_contents))
}))
```
