---
title: "Landings and logbooks: gears and depth classes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Landings and logbooks: gears and depth classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# pax joins tables by their shared columns, and dplyr says so ("Joining
# with ..."), and duckdb tells where it keeps its extensions; drop those
# messages and keep the others
local({
  hook <- knitr::knit_hooks$get("message")
  knitr::knit_hooks$set(message = function(x, options) {
    x <- strsplit(x, "\n")[[1]]
    x <- x[!grepl(
      "Joining with `by|^(#> )?$|duckdb|persists across|shared_home",
      x
    )]
    if (length(x) == 0) "" else hook(paste0(x, "\n", collapse = ""), options)
  })
})
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.5
)
```

The `landings` table of a pax database is the landings register by year,
month, gear, country and ICES area; the `logbook` table the compiled
logbooks by tow, with position, depth and effort. This vignette covers the
two things that most often go wrong with them: gears and depths. All data
here are simulated.

```{r, message = FALSE}
library(pax)
library(dplyr, warn.conflicts = FALSE)

pcon <- pax_connect()
```

## Gear codes

Both tables have the gear as an MFDB gear code (`mfdb_gear_code`: BMT bottom
trawl, LLN longline, DSE Danish seine, GIL gillnet, ...), mapped from the
MFRI gear code (`gear_id`, *veiðarfæri*) with `biota.gear_mapping`. The
mapping is in the package as `gear_mapping`:

```{r}
data("gear_mapping", package = "pax")
gear_mapping |> filter(gear_id %in% c(1, 2, 6, 9, 91, 92))
```

`biota.gear_mapping` has no MFDB code for gear 91 (monkfish gillnet), so
those rows have `mfdb_gear_code` NA in `pax_db`; tidypax mapped it to GIL.
Monkfish lost most of its commercial samples in 2004–2015 and about 14 kt of
landings moved to "other" (14-mon). `gear_mapping` adds gear 91 as GIL
(`source == "pax"`), and `pax_fill_mfdb_gear_code()` fills the missing codes
from it:

```{r}
landings <- expand.grid(
  year = 2021:2024,
  month = 1:12,
  gear_id = c(6, 1, 91, 9)
)
landings$species <- 1
landings$ices_area <- "5a"
landings$country <- "Iceland"
landings$boat_id <- landings$gear_id * 100 + landings$month %% 3
gm <- gear_mapping[gear_mapping$source == "biota.gear_mapping", ]
landings$mfdb_gear_code <- as.character(gm$mfdb_gear_code)[
  match(landings$gear_id, gm$gear_id)
]
landings$catch <- c(`6` = 3e5, `1` = 2e5, `91` = 5e4, `9` = 1e4)[
  as.character(landings$gear_id)
]
pax_import(pcon, landings)

tbl(pcon, "landings") |>
  count(gear_id, mfdb_gear_code) |>
  collect()

tbl(pcon, "landings") |>
  pax_fill_mfdb_gear_code() |>
  count(gear_id, mfdb_gear_code) |>
  collect()
```

Only missing codes are filled. 13-cas does this on its stations before the
length distributions; do it on the landings too before grouping by gear.

## Landings by gear group

`pax_landings_by_gear()` groups gears with a `gear_group` list. The names are
the groups, the values the MFDB codes in them; `pax_add_other()` marks the
default group for every gear not listed, and an `NA` in a group takes the
rows with no gear:

```{r}
gear_group <- list(
  BMT = c("BMT", "NPT", "SHT", "PGT"),
  LLN = "LLN",
  GIL = "GIL",
  Other = pax_add_other()
)
by_gear <- tbl(pcon, "landings") |>
  pax_fill_mfdb_gear_code() |>
  pax_landings_by_gear(gear_group = gear_group)

by_gear |>
  group_by(year, gear_name) |>
  summarise(catch_t = sum(catch, na.rm = TRUE) / 1e3, .groups = "drop") |>
  collect() |>
  tidyr::pivot_wider(names_from = gear_name, values_from = catch_t) |>
  arrange(year)
```

`catch` is in kg. Without the default group, landings with other gears get
no group: in `pax_si_scale_by_landings()` they are left out of the raising,
with a message. The same `gear_group` lists drive the age–length keys and
the advice and tech report figures (`hafroreports::hr_advice_plot_landings()`,
`hr_techreport_plot_landings_gear()`), so use the stock's own groups there.

The landings by fishing year (September to August, from 1991). Note that
`catch_kt` is `sum(catch) / 1000`, so it is in tonnes when `catch` is in kg,
as in the `landings` table:

```{r}
tbl(pcon, "landings") |>
  pax_landings_fishingyear_summary(ignore_final_year = FALSE) |>
  collect()
```

## Depth classes of logbook records

`pax_add_ocean_depth_class()` puts each record in a depth class from its
`ocean_depth` column. Records with no recorded depth get the mean depth of
the bathymetry around their own position: their h3 cells are matched to the
`ocean_depth` table at `fill_resolution` (default 6, cells of about 36 km²),
and at two coarser resolutions where that finds nothing. Records without a
position stay "Unknown". The bathymetry comes from
`pax_marmap_ocean_depth()` (a cached NOAA grid around Iceland, imported by
`pax_from_mar()`).

```{r}
pax_import(pcon, pax_marmap_ocean_depth())

# Tows at the centres of random gridcells (statistical subrectangles)
data("gridcell", package = "pax")
set.seed(3)
cells <- gridcell[sample(nrow(gridcell), 100, replace = TRUE), ]
logbook <- data.frame(
  year = rep(2021:2024, each = 25),
  month = sample(1:12, 100, replace = TRUE),
  species = 1,
  mfdb_gear_code = sample(c("BMT", "LLN"), 100, replace = TRUE),
  gridcell = cells$gridcell,
  lat = cells$lat,
  lon = cells$lon
)
logbook$ocean_depth <- round(runif(100, 40, 500))
logbook$ocean_depth[1:20] <- NA          # no depth recorded
logbook$lat[1:3] <- logbook$lon[1:3] <- NA  # and no position either
logbook$catch <- round(rlnorm(100, 7, 1))
logbook$tow_time <- ifelse(logbook$mfdb_gear_code == "BMT", 240, NA)
logbook$tow_hooks <- ifelse(logbook$mfdb_gear_code == "LLN", 20000, NA)
logbook$tow_num_nets <- NA_real_
logbook$catch_total <- logbook$catch * 2
pax_import(pcon, logbook)

tbl(pcon, "logbook") |>
  mutate(depth = ifelse(is.na(ocean_depth), "filled", "recorded")) |>
  pax_add_ocean_depth_class(breaks = c(0, 100, 200, 300)) |>
  count(depth, ocean_depth_class) |>
  collect() |>
  arrange(ocean_depth_class) |>
  tidyr::pivot_wider(names_from = depth, values_from = n, values_fill = 0)
```

`breaks` are the class bounds; depths beyond the last bound form a plus
group ("300+"). `hafroreports::hr_techreport_plot_catchdepth()` uses this for
the catch by depth figure.

**Pitfall:** before pax branch `smn-survey` (commit a56fe05), every missing
depth got one and the same value (the mean of the whole bathymetry table,
about 950 m), so all records without a depth landed in the deepest class.
07-bli and 60-norway-redfish work around it with their own depth classes
(`bli_depth_class()`, `sfv_depth_class()`), from the mean depth of the
gridcell. With a current pax the workaround is not needed, but the numbers
differ a little (cells around the position instead of the gridcell). Check
the share of catch in "Unknown" and in the classes of filled records.

## CPUE

`pax_add_cpue()` takes the effort from the first of tow time (hours), hooks
(thousands) or nets; Danish seine hauls count as one. Records with none of
these get `effort_na` (default 1, i.e. one hour); `NULL` leaves them out.

```{r}
tbl(pcon, "logbook") |>
  pax_add_cpue() |>
  pax_logbook_cpue_plot()
```

`hafroreports::hr_techreport_plot_cpue()` draws the same figure from a
`pax_db`.

```{r, include = FALSE}
DBI::dbDisconnect(pcon)
```

## AI use

This vignette was drafted with Claude (Anthropic) in October 2026 and has not
yet been checked by a person. (MFRI policy on AI use.)
