---
title: "Building and opening a pax database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building and opening a pax database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# duckdb tells where it keeps its extensions; drop that message and keep the
# others
local({
  hook <- knitr::knit_hooks$get("message")
  knitr::knit_hooks$set(message = function(x, options) {
    x <- strsplit(x, "\n")[[1]]
    x <- x[!grepl("^(#> )?$|duckdb|persists across|shared_home", x)]
    if (length(x) == 0) "" else hook(paste0(x, "\n", collapse = ""), options)
  })
})
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Every pax calculation runs on a local DuckDB database, the `pax_db` of a
stock repo. It is built once from the MFRI Oracle database (`mar`) with
`pax_from_mar()`, kept as a targets object, and everything downstream reads
only it. This vignette shows how it is built in the stock repos, what goes
in it, and how to open, extend and inspect one. The database parts need
`mar` and are not run here; the rest runs on a toy database.

## What `pax_connect()` needs

`pax_connect()` opens (or creates) a DuckDB file, or an in-memory database
by default, and loads two DuckDB extensions: `spatial` and the community
extension `h3`. The first time on a machine DuckDB downloads them from the
internet (to `~/.duckdb`); after that it works offline.

```{r}
library(pax)
library(dplyr, warn.conflicts = FALSE)

path <- tempfile(fileext = ".duckdb")
pcon <- pax_connect(path)
```

## Building from mar

`pax_from_mar()` connects to mar (VPN when working remotely) and imports the
standard tables for the given species and years: `station`, `measurement`,
`ldist`, `aldist`, `lw_coeffs`, `sampling`, `landings`, `logbook`,
`quotatransfer`, the fixed survey station lists `strata_stations`, the
bathymetry `ocean_depth` and the strata polygons (`old_strata`,
`new_strata_spring`, ...). The saithe call (03-sai):

```{r, eval = FALSE}
pax_db <- pax_from_mar(
  species = 3,
  year_start = 1980,
  year_end = 2026,
  # Commercial (1, 2, 4, 8), survey otoliths (10, 11), spring (30) and
  # autumn (35) surveys
  sampling_type = c(1, 2, 4, 8, 10, 11, 30, 35),
  # Landings since 1903, for the landings history of the tech report
  landings_year_start = 1903,
  dbdir = "pax.duckdb"
)
```

Other arguments the stock repos use:

* `ices_area_like = c("5a%", "14%")`: landings of more areas (07-bli, 08-usk);
* `skip_trips`: by default the stomach-sampling trips (`MAG%`, `MO%`) are
  left out of `station`, which tidypax kept; cod lost up to 26% of its
  commercial samples, see `extra_tables = "station_skipped"` below;
* `gridcell_from_position = TRUE`: stations without a rectangle get the one
  of their position (otherwise they get no region in the age–length keys);
* `logbook_species`, `logbook_year_start`: logbooks of other species or
  years than the stock's.

### Extra tables

Tables the stock repos used to read from mar directly are optional:

```{r}
pax_from_mar_extra_tables()
```

01-cod takes most of them (TAC left, catch outside the quotas, old landings,
stomach-trip samples, all samples for length–weight); 07-bli takes `sample`.
See `?pax_from_mar` for each.

### Foreign data

Data of other countries are never part of the package. They are read when
the database is built, either from tables in mar (`foreign_tables`):

```{r}
pax_foreign_mar_tables()$name
```

or from files delivered by the other institutes (`foreign_files`, a named
list of file type to path):

```{r}
pax_foreign_file_types()
```

The German (Walther Herwig) survey off Greenland is read from the
Thünen-Institut exports with `ger_survey_path` (a directory or files),
`ger_species` and `ger_species_code`, into the tables `ger_station`,
`ger_catch`, `ger_ldist` and `ger_strata` (see `?pax_ger_survey`). 05-reg and
08-usk:

```{r, eval = FALSE}
pax_from_mar(
  species, year_start, year_end,
  sampling_type = c(1, 2, 4, 8, 30, 34, 35),
  ices_area_like = c("5a%", "14%"),
  ger_survey_path = ger_survey_files,            # config.R, from an env var
  ger_species = "BROSME BROSME",
  ger_species_code = species,
  foreign_files = list(greenland_logbooks = path.expand(greenland_logbook_file))
)
```

All file paths are checked before the (slow) import from mar starts. Keep
the paths in `config.R`, set from environment variables, and never commit
the files. A `pax_db` built with foreign data holds those data: MFRI may
work with them but not share them, so a copy of the database must not leave
MFRI (see the 01-cod README).

## In a targets pipeline

The stock repos build `pax_db` as the first target of
`script_assessment_model.R`:

```{r, eval = FALSE}
tar_target(
  pax_db,
  if (nzchar(Sys.getenv("PAX_SOURCE_DB"))) {
    pax_connect(Sys.getenv("PAX_SOURCE_DB"))
  } else {
    pax_from_mar(species, year_start, year_end, ...)
  },
  format = pax_tar_format_duckdb()
)
```

* `pax_tar_format_duckdb()` writes the database into the targets store
  (copying an in-memory database to the file) and opens it **read-only** in
  every target that uses it; reading it doesn't change its hash.
* `PAX_SOURCE_DB` points at an existing `pax_db` file (e.g. a copy of
  `_assessment_model/objects/pax_db`). With it set, the pipeline runs
  without mar or the VPN; the repos check in `run.R` that no other code
  reads mar (`check_no_mar()`, e.g. 01-cod, 03-sai).
* Derived tables go in Parquet with `pax_tar_format_parquet()`, which drops
  the `geom` and `h3_cells` columns Parquet can't store.
* targets doesn't track package versions: after installing a new pax, run
  `targets::tar_invalidate(pax_db)` to rebuild it. Also check after
  `renv::restore()` that you got the pinned pax branch, e.g.
  `exists("pax_si_length_index", asNamespace("pax"))`.

The vignette "Incorporating external data" shows a live targets run with
extra imports.

## Importing and inspecting

`pax_import()` writes a data.frame, an sf object, a lazy query or a CSV file
into the database, and records where it came from. A table with `lat`/`lon`
(or `begin_lat`, `begin_lon`, `end_lat`, `end_lon`) gets a `geom` column and
its h3 cells, polygons get the cells they cover; that is what the strata and
depth joins use.

```{r}
tows <- data.frame(
  sample_id = 1:3,
  year = 2025,
  lat = c(64.1, 65.3, 66.6),
  lon = c(-23.5, -25.1, -18.2)
)
pax_import(pcon, tows, cite = "Toy tows, made up for this vignette")
pax_import(
  pcon,
  pax_decorate(
    data.frame(species = 1, a = 0.01, b = 3),
    cite = "Toy length-weight coefficients",
    name = "lw_coeffs"
  )
)
colnames(tbl(pcon, "tows"))
pax_contents(pcon) |> collect()
```

The table name defaults to the variable name, or the `name` given to
`pax_decorate()`, and the citation to the one given to `pax_decorate()` (the
`pax_mar_*` functions set both). Importing over an existing table needs
`overwrite = TRUE`.

The database is a plain DuckDB file, so it can be opened again later,
read-only as targets does:

```{r}
DBI::dbDisconnect(pcon)
pcon <- pax_connect(path, read_only = TRUE)
DBI::dbListTables(pcon)
tbl(pcon, "tows") |> select(sample_id, lat, lon) |> collect()
DBI::dbDisconnect(pcon)
```

## AI use

This vignette was drafted with Claude (Anthropic) in October 2026 and has not
yet been checked by a person. (MFRI policy on AI use.)
