---
title: "Survey indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survey indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# pax joins tables by their shared columns, and dplyr says so ("Joining
# with ..."), and duckdb tells where it keeps its extensions; drop those
# messages and keep the others
local({
  hook <- knitr::knit_hooks$get("message")
  knitr::knit_hooks$set(message = function(x, options) {
    x <- strsplit(x, "\n")[[1]]
    x <- x[!grepl(
      "Joining with `by|^(#> )?$|duckdb|persists across|shared_home",
      x
    )]
    if (length(x) == 0) "" else hook(paste0(x, "\n", collapse = ""), options)
  })
})
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A stratified survey index in pax takes four steps, each a function that
takes a lazy table (a dplyr query on the pax database) and returns another:

1. `pax_si_by_length()`: join the stations to the length distributions, and
   turn the counts into fish per square nautical mile (tow length × width)
   and biomass (length–weight relationship);
2. `pax_si_scale_by_strata()` or `pax_si_scale_by_strata_stations()`: give
   each station a stratum and multiply by the stratum area over the number of
   stations in the stratum;
3. `pax_si_strata_summary()`: sum within strata, for one length range, with
   standard deviations;
4. `pax_si_year_summary()`: sum over strata to the index and its CV by year.

`pax_si_length_index()` runs all four for a fixed station list and several
length ranges, as the category 3 stocks use. In the stock repos the at-age
indices for SAM come from `hafroreports::hr_input_data_si_index()`, which runs
the same chain with an age–length key (see the vignette on length
distributions and age–length keys).

This vignette builds a tiny database of simulated survey data and runs both.
Nothing here comes from real surveys; only the strata polygons are the
package's own.

## A toy survey

The stations sit at the centres of 12 statistical subrectangles (gridcells),
three in each of four strata of the spring survey (`new_strata_spring`).

```{r, message = FALSE}
library(pax)
library(dplyr, warn.conflicts = FALSE)

pcon <- pax_connect(tempfile(fileext = ".duckdb"))
```

```{r, results = "hide", warning = FALSE}
# Strata polygons (with stratum areas) shipped with pax
pax_import(pcon, pax_def_strata("new_strata_spring"))
```

```{r}
grid <- data.frame(
  gridcell = c(2741, 3213, 3223, 3202, 3703, 3714,
               3161, 3631, 3643, 6113, 6623, 7144),
  lat = c(62.875, 63.125, 63.125, 63.375, 63.625, 63.625,
          63.375, 63.875, 63.625, 66.125, 66.625, 67.125),
  lon = c(-24.75, -21.75, -22.75, -20.25, -20.75, -21.25,
          -16.75, -13.75, -14.75, -11.75, -12.75, -14.25)
)

set.seed(42)
station <- merge(grid, data.frame(year = 2021:2024))
station <- station[order(station$year, station$gridcell), ]
station$sample_id <- seq_len(nrow(station))
station$sampling_type <- 30          # spring survey
station$month <- 3
station$tow_number <- rep(1:12, 4)
station$gear_id <- 73
station$mfdb_gear_code <- "BMT"
# Station id as in mar: rectangle, tow number, gear
station$station <- (station$gridcell %/% 10) * 1e4 +
  station$tow_number * 100 + station$gear_id
station$tow_length <- round(runif(nrow(station), 3.5, 4.2), 2)
station$tow_depth <- round(runif(nrow(station), 100, 400))
station$fixed <- 1
station$begin_lat <- station$lat
station$begin_lon <- station$lon
station$end_lat <- station$lat + 0.05
station$end_lon <- station$lon
station$lat <- station$lon <- NULL
pax_import(pcon, station)

# Lengths (cm) of the fish of each tow: more and larger fish in later years
ldist <- do.call(rbind, lapply(seq_len(nrow(station)), function(i) {
  n <- rpois(1, 20 + 5 * (station$year[i] - 2021))
  if (n == 0) return(NULL)
  len <- round(rnorm(n, 35 + station$year[i] - 2021, 10))
  len <- table(len[len > 0])
  data.frame(
    sample_id = station$sample_id[i],
    species = 1,
    length = as.numeric(names(len)),
    sex = NA_real_,
    count = as.numeric(len)
  )
}))
pax_import(pcon, ldist)

# Length-weight coefficients (weight in g = a * length^b)
pax_import(pcon, data.frame(species = 1, a = 0.01, b = 3), name = "lw_coeffs")
```

In a database built by `pax_from_mar()` these tables (`station`, `ldist`,
`lw_coeffs`, the strata and `strata_stations`) are already there.

## Strata from tow positions

By default each tow goes in the stratum of the h3 cell of its first position
(the strata tables hold the h3 cells of each polygon):

```{r}
by_length <- tbl(pcon, "station") |>
  filter(sampling_type == 30, coalesce(tow_number, 0) %in% 0:35) |>
  pax_si_by_length()

index_pos <- by_length |>
  pax_si_scale_by_strata("new_strata_spring") |>
  pax_si_strata_summary(length_range = c(0, 500)) |>
  pax_si_year_summary() |>
  collect() |>
  arrange(year)
index_pos |> select(year, si_N, si_biomass, si_biomass_cv)
```

`si_abund` is in thousands of fish and `si_biomass` in tonnes; `si_N` is the
number of strata with fish. `pax_si_strata_summary()` is where the length
range is applied, and `pax_si_year_summary()` drops strata without fish.

## Strata from a fixed station list

The tidypax indices, and so the published ones, put each station in the
stratum of the fixed station list (`biota.strata_stations`), by station id.
`pax_from_mar()` imports the list as the table `strata_stations` (with the
gillnet survey's `smn_strata` list too). Here is a toy list that agrees with
the positions except for one station near a stratum boundary, which the list
puts in stratum 1:

```{r}
strata_stations <- by_length |>
  pax_si_scale_by_strata("new_strata_spring") |>
  ungroup() |>
  distinct(sampling_type, station, stratum) |>
  collect() |>
  mutate(stratification = "new_strata")
moved <- station$station[station$gridcell == 3202][1]
strata_stations$stratum[strata_stations$station == moved] <- 1
pax_import(pcon, strata_stations)
```

`pax_si_strata_stations()` reads the list of one stratification and adds the
stratum area of each survey (converted from km² to square nautical miles);
`pax_si_scale_by_strata_stations()` then uses it in place of
`pax_si_scale_by_strata()`:

```{r}
ss <- pax_si_strata_stations(
  pcon,
  stratification = "new_strata",
  area_tables = c(`30` = "new_strata_spring")
)
head(ss, 3)

index_list <- by_length |>
  pax_si_scale_by_strata_stations(ss |> filter(sampling_type == 30)) |>
  pax_si_strata_summary(length_range = c(0, 500)) |>
  pax_si_year_summary() |>
  collect()

full_join(
  index_pos |> select(year, by_position = si_biomass),
  index_list |> select(year, by_list = si_biomass),
  by = "year"
) |>
  arrange(year)
```

One station moved, and every year changes: the stratum it left and the one
it joined now have different numbers of stations. In the saithe spring survey
about 50 stations a year differ, which moved the index by −47% to +64% in
single years (03-sai). Use the station list when the published index did.

`pax_si_scale_by_strata(strata_tbl, strata_stations = ...)` also takes a
station list, but takes the areas from the one strata table; with
`pax_si_scale_by_strata_stations()` each survey keeps its own areas (spring
and autumn strata differ). The hafroreports equivalent is
`hr_si_scale_by_strata_stations()`, and `hr_input_data_si_index(strata_name =,
strata_stations =)` uses it, as in 03-sai and 01-cod:

```{r, eval = FALSE}
hafroreports::hr_input_data_si_index(
  pax_db,
  sampling_type = 30,
  tow_number = 0:35,
  strata_name = "old_strata",
  strata_stations = strata_stations |>
    dplyr::filter(sampling_type == 30, stratification == "old_strata")
)
```

## Indices by length range

The category 3 stocks (e.g. 04-whg, 12-rjr, 13-cas) use biomass indices of
length ranges with the fixed station list. `pax_si_length_index()` selects
the stations (survey, tows, gears, years, fixed stations) and returns one
index per length range:

```{r}
pax_si_length_index(
  pcon,
  ss,
  length_ranges = list(total = c(0, 500), large = c(40, 500)),
  sampling_type = 30,
  tow_number = 0:35
) |>
  select(length_range, year, srN, b, b_cv)
```

Length ranges are exclusive at both ends: `c(40, 500)` counts fish of 41 cm
and over (in whole cm), as tidypax did. A range written as "40 cm and over"
is `c(39, 500)`.

The other arguments mirror the stock repos:

* `tow_number = 0:35` (spring survey) or `0:75` (autumn survey); stations
  without a tow number count as tow 0;
* `gear_id = 77:78` for the autumn survey;
* `skip_years = 2011` for the autumn survey, which covered only part of the
  area that year;
* `fixed_only = TRUE` keeps the fixed stations (`fixed == 1`), as 04-whg
  does for the spring survey.

A year with stations but no fish in a length range has no row by default (as
tidypax). With `complete_years = TRUE` it gets an index of 0. This matters
for the rfb rule, which takes the last five rows of the index: a missing
year shifts the index ratio.

## Pitfalls

* **Tow filter.** `tow_number` is the tow of a survey station, but the haul
  number of a commercial sample. Never apply a survey tow filter to
  commercial samples (03-sai lost 8–14% of its samples a year that way).
* **Strata from positions vs the station list.** See above. Check which one
  the published index used.
* **Length ranges are exclusive.** `c(30, 500)` starts at 31 cm.
* **Gillnet survey (SMN, sampling type 34).** Stations are per net, 0.5 nm
  per net; use the `smn_strata` station list (06-lin, 08-usk).
* **Large hauls.** Old scripts scaled named hauls down (e.g. to 5%); use
  `hr_input_data_si_index(haul_scalar = ...)` for that, not
  `pax_si_scale_winsorize()`, which is a different, generic rule (and in
  this version of pax changes nothing: it takes the quantile of a column
  that doesn't exist, so no haul is above it).
* **The ldist table is already raised** to the counted fish at import, so
  don't apply `pax_ldist_scale_abund()` to it (see the vignette on length
  distributions).

```{r, include = FALSE}
DBI::dbDisconnect(pcon)
```

## AI use

This vignette was drafted with Claude (Anthropic) in October 2026 and has not
yet been checked by a person. (MFRI policy on AI use.)
